---
title: 'Exercises With plotly and State Energy Data System (SEDS) Data'
author: "Nathan Sowatskey - March 12 2017 - Coursera/John Hopkins Data Science Specilisation"
always_allow_html: yes
---

```{r load_libraries, echo = FALSE, warning = FALSE, message = FALSE}
require("tidyr")
require("plotly")
require("readxl")
```

#Introduction

These are exercises with plotly and data from the [State Energy Data System (SEDS)](https://www.eia.gov/state/seds/seds-data-complete.php).

#Data

The original data source is available on a state by state basis, or for the US as a whole, in a "wide" format. 

The code below downloads the data, imports it and tidies it into a long format. This code also includes unit codes, which, at the time of writing, are of my own invention.

The code is hidden in this file so that the plots can be seen, which is the focus of the coursework. The code can be seen in the Rmd file in this [Github repository]().

```{r setting_up_variables, echo = FALSE}
#All US states, excluding "US", including "DC"
us_states <- c(state.abb, "DC")
```

```{r download_data, echo = FALSE}
#The form of the URL for a data file is this URL with use_<state code>.csv
base_use_url <- "https://www.eia.gov/state/seds/sep_use/total/csv/"
#Save file locally for cache purposes
for (state in us_states) {

  file_name <- paste("use_", state, ".csv", sep = "")
  if (!file.exists(file_name)) {
    use_data_url <- paste(base_use_url, file_name, sep = "")
    #Using wget as the method here is important, as curl seems not to work (on OSX at least)
    download.file(use_data_url, file_name, method = "wget")
  }
}
```

```{r create_data_frames, echo = FALSE}
wide_col_names <- c("Data_Status", "State", "MSN", "1960", "1961", "1962", "1963", "1964", "1965", "1966", "1967", "1968", "1969", "1970", "1971", "1972", "1973", "1974", "1975", "1976", "1977", "1978", "1979", "1980", "1981", "1982", "1983", "1984", "1985", "1986", "1987", "1988", "1989", "1990", "1991", "1992", "1993", "1994", "1995", "1996", "1997", "1998", "1999", "2000", "2001", "2002", "2003", "2004", "2005", "2006", "2007", "2008", "2009", "2010", "2011", "2012", "2013", "2014")
#Create a dataframe of the required size to merge all data.
#One empty row allows me to set the column types.
all_states_wide_df <- as.data.frame(matrix(0, ncol = 58, nrow = 1))
colnames(all_states_wide_df) <- wide_col_names
#Change the first three columns to character, so that I can merge factor types in later
all_states_wide_df$Data_Status <- as.character("")
all_states_wide_df$State <- as.character("")
all_states_wide_df$MSN <- as.character("")

#It would be good if wide_col_classes could be used, but the numbers are quoted, so this sees not to work. This note is simply to make it clear that I tried.
#The result is correct anyway, but probably using more memory than necessary.
wide_col_classes <- c("factor", "factor", "factor", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric")

for (state in us_states) {
  file_name <- paste("use_", state, ".csv", sep = "")
  #check.names = FALSE stops the 'X' appearing in front of the name.
  all_states_wide_df <- rbind(all_states_wide_df, read.csv(file_name, col.names = wide_col_names, check.names = FALSE))
}
#Delete first, blank, row.
all_states_wide_df <- all_states_wide_df[-1,]
#Save the Data_Status value, and then drop that column as we don't need it.
data_status <- all_states_wide_df[1,]$Data_Status
all_states_wide_df$Data_Status <- NULL
#Now change to factors
all_states_wide_df$State <- as.factor(all_states_wide_df$State)
all_states_wide_df$MSN <- as.factor(all_states_wide_df$MSN)
```

```{r tidy_data_frame, echo = FALSE}
#Create a long format, tidy, dataframe
all_states_long_df <- gather(data = all_states_wide_df, key = "Year", value = "value", -State, -MSN)
#Some sanity checking
stopifnot((all_states_wide_df[which(all_states_wide_df$State == "US" & all_states_wide_df$MSN == "ARICB"), 3] == all_states_long_df[which(all_states_long_df$State == "US" & all_states_long_df$MSN == "ARICB" & all_states_long_df$Year == "1960"), 4]))
```

```{r MSN_units, echo = FALSE}
#The meaning of the MSN codes is in this file:
if (!file.exists("Codes_and_Descriptions.xlsx")) {
  download.file("https://www.eia.gov/state/seds/CDF/Codes_and_Descriptions.xlsx", "Codes_and_Descriptions.xlsx", method = "wget")
}

msn_codes <- read_excel("./Codes_and_Descriptions.xlsx", sheet = 2, col_names = FALSE, skip = 8)
msn_codes <- msn_codes[, 2:4]
msn_codes <- na.omit(msn_codes)
colnames(msn_codes) <- c("MSN", "Description", "Units_Desc")
msn_codes$MSN <- as.factor(msn_codes$MSN)
#Create easier to use abbreviations for units. I struggled to find a syntactically elegant way to do this ...
#There is also a question in my mind of whether there are already standard/common short codes for these units. 
#It seems odd that, apparently, there are not. I have asked the data provider, so this may be updated.
msn_codes$Units_Code <- as.character("") 
for (row in 1:nrow(msn_codes)){
  if (msn_codes[row,]$Units_Desc == "Billion Btu") {
    msn_codes[row,]$Units_Code <- "BBtu"
  }
  if (msn_codes[row,]$Units_Desc == "Million Btu") {
    msn_codes[row,]$Units_Code <- "MBtu"
  }
  if (msn_codes[row,]$Units_Desc == "Thousand barrels") {
    msn_codes[row,]$Units_Code <- "Kbrls"
   }
  if (msn_codes[row,]$Units_Desc == "Dollars per million Btu") {
    msn_codes[row,]$Units_Code <- "$/MBtu"
  }
  if (msn_codes[row,]$Units_Desc == "Million dollars") {
    msn_codes[row,]$Units_Code <- "M$"
  }
  if (msn_codes[row,]$Units_Desc == "Thousand short tons") {
    msn_codes[row,]$Units_Code <- "KShrtTon"
  }
  if (msn_codes[row,]$Units_Desc == "Million Btu per short ton") {
    msn_codes[row,]$Units_Code <- "MBtu/ShrtTon"
  }
  if (msn_codes[row,]$Units_Desc == "Million Btu per barrel") {
    msn_codes[row,]$Units_Code <- "MBtu/Brl"
  }
  if (msn_codes[row,]$Units_Desc == "Million kilowatthours") {
    msn_codes[row,]$Units_Code <- "Mkwh"
  }
  if (msn_codes[row,]$Units_Desc == "Thousand Btu per kilowatthour") {
    msn_codes[row,]$Units_Code <- "KBtu/kwh"
  }
  if (msn_codes[row,]$Units_Desc == "Percent share") {
    msn_codes[row,]$Units_Code <- "%shr"
  }
  if (msn_codes[row,]$Units_Desc == "Million chained (2009) dollars") {
    msn_codes[row,]$Units_Code <- "M$chnd"
  }
  if (msn_codes[row,]$Units_Desc == "Thousand Btu per chained (2009) dollar") {
    msn_codes[row,]$Units_Code <- "KBtu/$chnd"
  }
  if (msn_codes[row,]$Units_Desc == "Dollars") {
    msn_codes[row,]$Units_Code <- "$"
  }
  if (msn_codes[row,]$Units_Desc == "Percent") {
    msn_codes[row,]$Units_Code <- "%"
  }
  if (msn_codes[row,]$Units_Desc == "Million cubic feet") {
    msn_codes[row,]$Units_Code <- "Mcf"
  }
  #I am guessing that these two are the same, so they have the same short code.
  if (msn_codes[row,]$Units_Desc == "Thousand Btu per cubic foot") {
    msn_codes[row,]$Units_Code <- "KBtu/cf"
  }
  if (msn_codes[row,]$Units_Desc == "Thousand Btu per cubic feet") {
    msn_codes[row,]$Units_Code <- "KBtu/cf"
  }
  if (msn_codes[row,]$Units_Desc == "Thousand") {
    msn_codes[row,]$Units_Code <- "K"
  }
  if (msn_codes[row,]$Units_Desc == "Thousand cords") {
    msn_codes[row,]$Units_Code <- "Kcords"
  }
}
#Check there are no missing codes
stopifnot(nrow(msn_codes[msn_codes$Units_Code == "",]) == 0)
#Add units columns
all_states_long_df$Units_Code <- msn_codes$Units_Code[match(all_states_long_df$MSN, msn_codes$MSN)]
```

```{r subset_by_msn, echo = FALSE}
#Subset data by MSN so that I can plot for MSN
#Note that creating the matrix with a list data type is important, else you can't add dataframes into cells
msn_subsets_mtrx <- matrix(list(), nrow = 2, ncol = nrow(msn_codes))
col_counter <- 1
for (msn_code in msn_codes$MSN) {
  msn_subsets_mtrx[[1, col_counter]] <- msn_code
  msn_subsets_mtrx[[2, col_counter]] <- subset(all_states_long_df, MSN == msn_code)
  col_counter <- col_counter + 1
}
```

```{r plotly_lines}
#This syntax for using ployly in a loop comes from here:
#https://github.com/ropensci/plotly/issues/273
line_plot_gatherer <- htmltools::tagList()
col_counter <- 1
for (msn_code in msn_codes$MSN) {
  plot_df <- msn_subsets_mtrx[[2, col_counter]]
  if (nrow(plot_df) > 0) {
    line_plot_gatherer[[col_counter]] <- plot_ly(plot_df, x = ~Year, y = ~value, color = ~State, type = "scatter", mode = "lines") %>% 
      layout(title = paste("All US States -", msn_codes[msn_codes$MSN == msn_code, 2], "-", msn_codes[msn_codes$MSN == msn_code, 4]))
  }
  col_counter <- col_counter + 1
  #Stop before too many are created, just to save time and space  
  if (col_counter > 10) {
    break
  }
}
line_plot_gatherer
```


## Choropleth Maps
There is a bug with, possibly, plotting multiple maps that this code illustrates. 
This could be user error also.
It may be related to:
https://github.com/ropensci/plotly/issues/273
There are two main issues:
1 - Each map seems to be using the same dataframe in the map values, though the debug title seems to show that there is a distinct data frame when the map is created.
2 - Trying to plot more than two maps means that no data at all is shown in the maps.

I am keeping this here as an example I can refer to if needed, so I want to publish it also. The coursework deadline is today.

```{r, echo=FALSE, message=FALSE}
# Make state borders red
borders <- list(color = toRGB("red"))
# Set up some mapping options
map_options <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white')
)
map_plot_gatherer <- htmltools::tagList()
col_counter <- 1
for (msn_code in msn_codes$MSN) {
  plot_df <- msn_subsets_mtrx[[2, col_counter]]
  #Select a specific year
  year <- "2000"
  plot_df <- subset(plot_df, Year == year)
  if (nrow(plot_df) > 0) {
    map_plot_gatherer[[col_counter]] <- plot_ly(z = ~plot_df$value, text = ~plot_df$value, locations = ~plot_df$State, type = 'choropleth', locationmode = 'USA-states', color = plot_df$Value, colors = 'Blues', marker = list(line = borders)) %>%
    layout(title = paste("Debug sum value is -", sum(subset(plot_df, MSN == msn_code)$value)), geo = map_options)
  #The above is a debug title to see whether I realy am using different data sets
  #This is what the title should be:
  #layout(title = paste(msn_codes[msn_codes$MSN == msn_code, 2], "-", year, "-", msn_codes[msn_codes$MSN == msn_code, 4]), geo = map_options)
  }
  col_counter <- col_counter + 1
  #If more than two maps are created, then there is no data, just blank maps. See:
  #https://github.com/ropensci/plotly/issues/273
  #Stopping when col_counter > 2 works, to show two maps. Trying to show three did not work.
  if (col_counter > 3) {
    break
  }
}
map_plot_gatherer
```

## More Resources

- [The Plolty Website](https://plot.ly/)
- [The Plotly R API](https://plot.ly/r/)
- [The Plotly R Package on GitHub](https://github.com/ropensci/plotly)
- [The Plotly R Cheatsheet](https://images.plot.ly/plotly-documentation/images/r_cheat_sheet.pdf)
- ["Plotly for R" book by Carson Sievert](https://cpsievert.github.io/plotly_book/)