#State Energy Data Systems (SEDS) data for all US states, including DC, from 1960 to 2014F

# Context 

This dataset is derived from my general interest in energy systems. It was originally composed for [this exercise](https://data-science-projects.github.io/plotly_examples/seds/seds.html), as part of this [Coursera/John Hopkins Data Science Specilisation](https://www.coursera.org/specializations/jhu-data-science).

The code that produced this dataset is included below also.

# Content

The data is a composition of the State Energy Data Systems (SEDS) data for all US states, including DC, from 2016 to 2014F, for data released June 29, 2016. It has been tidied from a wide format to a long format, and includes unit codes for the values associated with the observations for each MSN code for each state for each year. 

The "F" in the final year number indicates that these are the final observations. There is a lag of some 18 months after year end and final readings.

The columns are:

 - state - State postal code, composed from the function `states.abb` and including "DC".
 - msn - A mnemonic series name identifying the value being observed.
 - year - Year of the observation.
 - value - Of the observation.
 - units_code, representing the units of the value, e.g. BBtu is Billion British Thermal Units.
 
Note that the units_codes are mostly my own invention, based on the [EIA Writng Style Guide](https://www.eia.gov/about/styleguide2015.pdf).

# Acknowledgements

Thank you to the [US Energy Information Administration](https://www.eia.gov) for making the data available.

Special thanks to Yvonne Taylor for guidance on style for the codes.

# Inspiration

The first goal for this data was to support some plotting and forecast testing exercises, which is a work in progress. To what extent do past observations predict future observations? Since the data is readily available, and consistent, within limits, over a long period, this format is a good basis for experimenting with techniques in that space. 

#Data Wrangling Code

The original data source is available on a state by state basis, or for the US as a whole, in a "wide" format. 

The code below downloads the data, imports it and tidies it into a long format. This code also includes unit codes, which, at the time of writing, are of my own invention.


```{r load_libraries, echo = FALSE, warning = FALSE, message = FALSE}
require("tidyr")
require("plotly")
require("readxl")
```

```{r setting_up_variables}
#All US states, excluding "US", including "DC"
us_states <- c(state.abb, "DC")
```

```{r download_data}
#The form of the URL for a data file is this URL with use_<state code>.csv
base_use_url <- "https://www.eia.gov/state/seds/sep_use/total/csv/"
#Save file locally for cache purposes
for (state in us_states) {
  file_name <- paste("use_", state, ".csv", sep = "")
  if (!file.exists(file_name)) {
    use_data_url <- paste(base_use_url, file_name, sep = "")
    #Using wget as the method here is important, as curl seems not to work (on OSX at least)
    download.file(use_data_url, file_name, method = "wget")
  }
}
```

```{r create_data_frames}
wide_col_names <- c("data_status", "state", "msn", "1960", "1961", "1962", "1963", "1964", "1965", "1966", "1967", "1968", "1969", "1970", "1971", "1972", "1973", "1974", "1975", "1976", "1977", "1978", "1979", "1980", "1981", "1982", "1983", "1984", "1985", "1986", "1987", "1988", "1989", "1990", "1991", "1992", "1993", "1994", "1995", "1996", "1997", "1998", "1999", "2000", "2001", "2002", "2003", "2004", "2005", "2006", "2007", "2008", "2009", "2010", "2011", "2012", "2013", "2014")
#Create a dataframe of the required size to merge all data.
#One empty row allows me to set the column types.
all_states_wide_df <- as.data.frame(matrix(0, ncol = 58, nrow = 1))
colnames(all_states_wide_df) <- wide_col_names
#Change the first three columns to character, so that I can merge factor types in later
all_states_wide_df$data_status <- as.character("")
all_states_wide_df$state <- as.character("")
all_states_wide_df$msn <- as.character("")

#It would be good if wide_col_classes could be used, but the numbers are quoted, so this sees not to work. This note is simply to make it clear that I tried.
#The result is correct anyway, but probably using more memory than necessary.
wide_col_classes <- c("factor", "factor", "factor", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric")

for (state in us_states) {
  file_name <- paste("use_", state, ".csv", sep = "")
  #check.names = FALSE stops the 'X' appearing in front of the name.
  all_states_wide_df <- rbind(all_states_wide_df, read.csv(file_name, col.names = wide_col_names, check.names = FALSE))
}
#Delete first, blank, row.
all_states_wide_df <- all_states_wide_df[-1,]
#Save the data_status value, and then drop that column as we don't need it.
data_status <- all_states_wide_df[1,]$data_status
all_states_wide_df$data_status <- NULL
#Now change to factors
all_states_wide_df$state <- as.factor(all_states_wide_df$state)
all_states_wide_df$msn <- as.factor(all_states_wide_df$msn)
```

```{r tidy_data_frame}
#Create a long format, tidy, dataframe
all_states_long_df <- gather(data = all_states_wide_df, key = "year", value = "value", -state, -msn)
#Some sanity checking
stopifnot((all_states_wide_df[which(all_states_wide_df$state == "US" & all_states_wide_df$msn == "ARICB"), 3] == all_states_long_df[which(all_states_long_df$state == "US" & all_states_long_df$msn == "ARICB" & all_states_long_df$year == "1960"), 4]))
```

```{r msn_units}
#The meaning of the MSN codes is in this file:
if (!file.exists("Codes_and_Descriptions.xlsx")) {
  download.file("https://www.eia.gov/state/seds/CDF/Codes_and_Descriptions.xlsx", "Codes_and_Descriptions.xlsx", method = "wget")
}

msn_codes <- read_excel("./Codes_and_Descriptions.xlsx", sheet = 2, col_names = FALSE, skip = 8)
msn_codes <- msn_codes[, 2:4]
msn_codes <- na.omit(msn_codes)
colnames(msn_codes) <- c("msn", "description", "units_desc")
msn_codes$msn <- as.factor(msn_codes$msn)
#Create easier to use abbreviations for units. I struggled to find a syntactically elegant way to do this ...
#There is also a question in my mind of whether there are already standard/common short codes for these units. 
#It seems odd that, apparently, there are not. I have asked the data provider, so this may be updated.
msn_codes$units_code <- as.character("") 
for (row in 1:nrow(msn_codes)){
  if (msn_codes[row,]$units_desc == "Billion Btu") {
    msn_codes[row,]$units_code <- "BBtu"
  }
  if (msn_codes[row,]$units_desc == "Million Btu") {
    msn_codes[row,]$units_code <- "MMBtu"
  }
  if (msn_codes[row,]$units_desc == "Thousand barrels") {
    msn_codes[row,]$units_code <- "Kbs"
   }
  if (msn_codes[row,]$units_desc == "Dollars per million Btu") {
    msn_codes[row,]$units_code <- "$/MMBtu"
  }
  if (msn_codes[row,]$units_desc == "Million dollars") {
    msn_codes[row,]$units_code <- "MM$"
  }
  if (msn_codes[row,]$units_desc == "Thousand short tons") {
    msn_codes[row,]$units_code <- "Kst"
  }
  if (msn_codes[row,]$units_desc == "Million Btu per short ton") {
    msn_codes[row,]$units_code <- "MMBtu/st"
  }
  if (msn_codes[row,]$units_desc == "Million Btu per barrel") {
    msn_codes[row,]$units_code <- "MMBtu/b"
  }
  if (msn_codes[row,]$units_desc == "Million kilowatthours") {
    msn_codes[row,]$units_code <- "MMkwh"
  }
  if (msn_codes[row,]$units_desc == "Thousand Btu per kilowatthour") {
    msn_codes[row,]$units_code <- "KBtu/kwh"
  }
  if (msn_codes[row,]$units_desc == "Percent share") {
    msn_codes[row,]$units_code <- "%shr"
  }
  if (msn_codes[row,]$units_desc == "Million chained (2009) dollars") {
    msn_codes[row,]$units_code <- "MM$chnd"
  }
  if (msn_codes[row,]$units_desc == "Thousand Btu per chained (2009) dollar") {
    msn_codes[row,]$units_code <- "KBtu/$chnd"
  }
  if (msn_codes[row,]$units_desc == "Dollars") {
    msn_codes[row,]$units_code <- "$"
  }
  if (msn_codes[row,]$units_desc == "Percent") {
    msn_codes[row,]$units_code <- "%"
  }
  if (msn_codes[row,]$units_desc == "Million cubic feet") {
    msn_codes[row,]$units_code <- "MMcf"
  }
  #I am guessing that these two are the same, so they have the same short code.
  if (msn_codes[row,]$units_desc == "Thousand Btu per cubic foot") {
    msn_codes[row,]$units_code <- "KBtu/cf"
  }
  if (msn_codes[row,]$units_desc == "Thousand Btu per cubic feet") {
    msn_codes[row,]$units_code <- "KBtu/cf"
  }
  if (msn_codes[row,]$units_desc == "Thousand") {
    msn_codes[row,]$units_code <- "K"
  }
  if (msn_codes[row,]$units_desc == "Thousand cords") {
    msn_codes[row,]$units_code <- "Kcords"
  }
}
#Check there are no missing codes
stopifnot(nrow(msn_codes[msn_codes$units_code == "",]) == 0)
#Add units columns
all_states_long_df$units_code <- msn_codes$units_code[match(all_states_long_df$msn, msn_codes$msn)]
```

```{r export_data}
write.csv(all_states_long_df, "seds_all_states_long.csv", row.names = FALSE)
```